\documentclass[12pt,a4paper]{article}
\usepackage{fontspec}
\setmainfont{Latin Modern Roman}
\usepackage{geometry}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{enumitem}
\usepackage{parskip}
\usepackage{titlesec}
\usepackage{array}

\geometry{margin=2.5cm}

\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    urlcolor=blue,
    citecolor=blue
}

\lstset{
    basicstyle=\ttfamily\small,
    breaklines=true,
    frame=single,
    backgroundcolor=\color{gray!10},
    columns=fullflexible
}

\title{
    \textbf{CENG 467 --- Term Project Progress Report} \\[0.5em]
    \Large Alignment via Direct Preference Optimization (DPO)
}

\author{
    Enes Ergün Hoşgör (300201053) \quad
    Samet Buldanlıoğlu (300201085) \quad
    Uğur Mert Çavuşoğlu (300201087)
}

\date{
    İzmir Institute of Technology (İYTE) \\
    CENG 467 Natural Language Understanding and Generation \\
    Prof.\ Dr.\ Aytuğ ONAN \quad Spring 2026 \\[0.5em]
    \textit{Submitted: May 14, 2026}
}

\begin{document}

\maketitle
\tableofcontents
\newpage

% ─────────────────────────────────────────────────────────────
\section{Project Title, Group Members, and Problem Definition}

\subsection*{Task Description}

This project addresses the task of \textbf{language model alignment}: steering a
pre-trained large language model (LLM) to produce outputs that are helpful, harmless,
and honest (HHH). Given a multi-turn conversation prompt, the system must generate a
response that a human annotator would prefer over a harmful or evasive alternative.

\subsection*{Why This Task Is Important}

LLMs trained on internet-scale corpora tend to reproduce harmful, biased, or unhelpful
content present in their training data. Classical alignment via Reinforcement Learning
from Human Feedback (RLHF) with Proximal Policy Optimization (PPO) requires a separate
reward model and is notoriously unstable to train. \textbf{Direct Preference
Optimization (DPO)} \cite{rafailov2023} eliminates the reward model entirely by
reformulating the RLHF objective as a supervised binary classification over
preference pairs, yielding a simpler, more stable training procedure.

\subsection*{Input and Output}

\begin{itemize}[noitemsep]
    \item \textbf{Input:} A multi-turn conversation prompt in the format
          \texttt{\textbackslash n\textbackslash nHuman: ...\textbackslash n\textbackslash nAssistant:}
    \item \textbf{Output:} A natural language assistant response
    \item \textbf{Supervision signal:} Pairwise preference labels
          (\texttt{chosen} vs.\ \texttt{rejected} response)
\end{itemize}

\subsection*{Task Type}

Conditional text generation with alignment. The task requires the model to learn a
policy that maximises the probability of preferred responses relative to a frozen
reference policy, without explicit reward signal.

\subsection*{Technical Challenge}

The core challenge is teaching the model to prefer safe, helpful responses over
harmful or evasive ones without catastrophic forgetting of language modeling ability.
DPO optimizes preference margins rather than pure likelihood, which can cause
perplexity to increase relative to a supervised fine-tuning (SFT) baseline when
training data is limited. Balancing the $\beta$ temperature parameter --- which
controls deviation from the reference policy --- is therefore critical.

% ─────────────────────────────────────────────────────────────
\section{Dataset / Benchmark Status}

\subsection*{Dataset Information}

\begin{table}[h]
\centering
\begin{tabular}{ll}
\toprule
\textbf{Property} & \textbf{Value} \\
\midrule
Name              & Anthropic HH-RLHF \cite{bai2022} \\
Source            & \url{https://huggingface.co/datasets/Anthropic/hh-rlhf} \\
Train split       & 160,800 examples (after filtering: 159,876) \\
Test split        & 8,505 examples \\
Input format      & Full conversation string (\texttt{chosen} and \texttt{rejected} fields) \\
Output format     & Preferred / dis-preferred assistant response \\
Answer types      & Free-form natural language \\
\bottomrule
\end{tabular}
\caption{HH-RLHF dataset properties}
\end{table}

\subsection*{Current Status}

The dataset has been fully downloaded, preprocessed, and verified locally and on
Google Colab / Kaggle. The preprocessing pipeline extracts structured
\texttt{prompt}, \texttt{chosen}, and \texttt{rejected} fields from the raw
conversation strings.

\subsection*{Preprocessing Steps Completed}

\begin{itemize}[noitemsep]
    \item Split each raw conversation on the last \texttt{\textbackslash n\textbackslash nAssistant:}
          occurrence to separate prompt from response
    \item Filtered malformed samples (empty responses, identical chosen/rejected pairs)
    \item Verified field lengths and encoding on a random 1,000-sample subset
\end{itemize}

The preprocessing function is:

\begin{lstlisting}[language=Python]
def extract_prompt_and_response(sample):
    prompt = sample["chosen"].rsplit("\n\nAssistant:", 1)[0] + "\n\nAssistant:"
    chosen = sample["chosen"].rsplit("\n\nAssistant:", 1)[1].strip()
    rejected = sample["rejected"].rsplit("\n\nAssistant:", 1)[1].strip()
    return {"prompt": prompt, "chosen": chosen, "rejected": rejected}
\end{lstlisting}

\subsection*{Known Issues}

\begin{itemize}[noitemsep]
    \item No official validation split; the test split is used for evaluation.
    \item Some long conversations are truncated at 512 tokens during model encoding.
    \item Current training uses 3--5k samples; full 50k+ scale is planned for the
          final submission.
\end{itemize}

% ─────────────────────────────────────────────────────────────
\section{Literature Review Progress}

\paragraph{Rafailov et al.\ (2023) --- DPO \cite{rafailov2023}.}
The foundational paper for this project. DPO shows that the optimal RLHF policy can
be expressed in closed form, allowing alignment to be solved as a binary
cross-entropy problem over preference pairs. The DPO loss is:
\[
\mathcal{L}_\text{DPO}(\pi_\theta) = -\mathbb{E}_{(x,y_w,y_l)} \left[
    \log \sigma\!\left(
        \beta \log \frac{\pi_\theta(y_w|x)}{\pi_\text{ref}(y_w|x)}
        - \beta \log \frac{\pi_\theta(y_l|x)}{\pi_\text{ref}(y_l|x)}
    \right)
\right]
\]
where $y_w$ is the chosen response, $y_l$ the rejected response,
$\pi_\text{ref}$ the frozen reference model, and $\beta$ controls deviation
from the reference policy.

\paragraph{Bai et al.\ (2022) --- HH-RLHF \cite{bai2022}.}
Introduces the dataset we use and the HHH (helpful, harmless, honest) framework.
Establishes the preference data collection methodology used by Anthropic and
provides baseline RLHF-PPO numbers that serve as our performance reference.

\paragraph{Ouyang et al.\ (2022) --- InstructGPT \cite{ouyang2022}.}
The seminal RLHF-PPO alignment paper. Shows significant improvement in helpfulness
and safety over base GPT-3. We use this as the upper-bound reference for
understanding the problem that DPO aims to simplify.

\paragraph{Hu et al.\ (2021) --- LoRA \cite{hu2021}.}
Low-Rank Adaptation enables fine-tuning of large models by injecting trainable
rank decomposition matrices into frozen weights. We use LoRA (rank=16) throughout
to fit TinyLlama-1.1B within the memory constraints of a single T4/A100 GPU.

\subsection*{Technical Direction}

Based on the reviewed literature, we follow the standard DPO pipeline:
(1) SFT on chosen responses to obtain a reference model,
(2) DPO fine-tuning using the SFT checkpoint as the frozen reference policy,
(3) evaluation with automatic metrics (perplexity, BERTScore) and qualitative
analysis on harmful prompts.

% ─────────────────────────────────────────────────────────────
\section{Baseline Models}

\subsection*{Baseline 1: Zero-Shot (Unaligned TinyLlama)}

The pretrained \texttt{TinyLlama/TinyLlama-1.1B-Chat-v1.0} model with no fine-tuning.
This represents the unaligned lower bound: the model responds to harmful prompts
without refusal and exhibits repetitive generation patterns.

\textbf{Status: Fully implemented and evaluated.}

\subsection*{Baseline 2: SFT (Supervised Fine-Tuning)}

TinyLlama-1.1B fine-tuned with LoRA (r=16) on the \texttt{chosen} responses only,
using supervised next-token prediction via \texttt{TRL SFTTrainer}. No preference
optimization is applied. Trained for 2 epochs on 5,000 samples.

\textbf{Status: Fully implemented and evaluated.}

\subsection*{Why These Baselines}

Zero-shot TinyLlama establishes the unaligned lower bound and isolates the effect of
any fine-tuning. SFT isolates the contribution of preference optimization: by training
only on chosen responses without DPO, any additional gains from the DPO model can be
attributed specifically to the contrastive preference objective rather than to
supervised exposure to high-quality responses.

% ─────────────────────────────────────────────────────────────
\section{Initial Experimental Results}

Table~\ref{tab:results} reports evaluation results on 100 test samples using two
automatic metrics:

\begin{itemize}[noitemsep]
    \item \textbf{Perplexity:} $\exp(\text{mean NLL loss})$ on chosen test responses.
          Lower is better.
    \item \textbf{BERTScore F1:} Semantic similarity between generated responses and
          chosen references using RoBERTa-Large. Higher is better.
\end{itemize}

\begin{table}[h]
\centering
\caption{Evaluation results on HH-RLHF test set (100 samples)}
\label{tab:results}
\begin{tabular}{llllp{2.8cm}}
\toprule
\textbf{Model} & \textbf{Perplexity} $\downarrow$ & \textbf{BERTScore F1} $\uparrow$ & \textbf{Status} \\
\midrule
Zero-Shot (TinyLlama-1.1B)   & 17.28 & 0.8303 & Complete \\
SFT (5k samples, 2 epochs)   & \textbf{10.08} & 0.8234 & Complete \\
DPO ($\beta$=0.1, 3k, 1 ep.) & 17.67 & \textbf{0.8272} & Complete \\
\bottomrule
\end{tabular}
\end{table}

\subsection*{Interpretation}

SFT achieves the lowest perplexity (10.08), significantly improving language modeling
quality by training on preferred responses. DPO shows slightly higher perplexity than
zero-shot due to limited training data (3,000 samples, 1 epoch) and the fact that
DPO optimizes for preference margins rather than pure language modeling likelihood.
However, DPO achieves the highest BERTScore F1 (0.8272), indicating better semantic
alignment with human-preferred responses despite the perplexity regression.

Qualitatively, the most important difference is behavior on harmful prompts. The
zero-shot model provides lock-picking instructions and endorses unsafe behavior,
while the DPO model shows greater tendency toward refusal. Full qualitative
comparison will be provided in the final report.

% ─────────────────────────────────────────────────────────────
\section{Planned Improvements and Technical Direction}

The core contribution of this project is scaling DPO training and rigorously evaluating
alignment quality. The planned pipeline proceeds as follows:

\begin{enumerate}[noitemsep]
    \item \textbf{Scale training data:} Increase from 5k/3k to 50k+ samples for both
          SFT and DPO to allow better preference generalization.
    \item \textbf{More training epochs:} 2--3 epochs for DPO to allow the preference
          objective to fully converge.
    \item \textbf{Win-rate evaluation:} Use a reward model to compute DPO vs.\ SFT
          win-rate on a held-out preference test set.
    \item \textbf{Ablation study:} Sweep $\beta \in \{0.01, 0.05, 0.1, 0.2, 0.5\}$
          and LoRA rank $\in \{8, 16, 32\}$ to identify optimal hyperparameters.
\end{enumerate}

% ─────────────────────────────────────────────────────────────
\section{Ablation / Prompt Sensitivity Plan}

We plan the following ablations to isolate the contribution of each design choice:

\begin{table}[h]
\centering
\begin{tabular}{ll}
\toprule
\textbf{Ablation} & \textbf{Variable} \\
\midrule
A1 & Zero-shot vs.\ SFT vs.\ DPO: effect of each training stage \\
A2 & $\beta \in \{0.01, 0.05, 0.1, 0.2, 0.5\}$: DPO temperature sensitivity \\
A3 & LoRA rank $\in \{8, 16, 32\}$: parameter efficiency trade-off \\
A4 & Learning rate $\in \{1\text{e-}5, 5\text{e-}5, 1\text{e-}4\}$: optimizer sensitivity \\
A5 & Training data scale: 5k vs.\ 20k vs.\ 50k samples \\
\bottomrule
\end{tabular}
\caption{Planned ablation study}
\end{table}

Results will be reported as a table of perplexity and BERTScore F1 for each
configuration, with the win-rate metric added once the reward model evaluation
is implemented.

% ─────────────────────────────────────────────────────────────
\section{Error Analysis and Generation Quality Plan}

\subsection*{Planned Error Analysis}

We will select 20--30 test samples where the DPO model fails to outperform the SFT
baseline and categorize failures as follows:

\begin{itemize}[noitemsep]
    \item \textbf{Hallucination:} Factually incorrect claims stated with confidence.
    \item \textbf{Refusal failure:} Model complies with a clearly harmful request.
    \item \textbf{Repetition:} Degenerate looping behavior (observed in SFT outputs).
    \item \textbf{Off-topic:} Response does not address the prompt.
\end{itemize}

\subsection*{Generation Quality Plan}

Once the full-scale model is trained, we will evaluate on a random sample of 50 test
prompts: factual correctness relative to gold chosen responses, fluency and coherence
of generated text, and hallucination rate (answers not grounded in any retrieved
evidence). We will include side-by-side qualitative examples of zero-shot, SFT, and
DPO outputs on a representative set of harmful, ambiguous, and benign prompts.

% ─────────────────────────────────────────────────────────────
\section{Ethical Considerations and Bias}

This project uses the Anthropic HH-RLHF dataset derived from human preference
annotations. Potential concerns include:

\begin{itemize}[noitemsep]
    \item \textbf{Annotator bias:} Human preference labels reflect the demographics
          and values of Anthropic's crowd workers, which may not generalize across
          cultures or languages.
    \item \textbf{Over-refusal:} DPO may cause the model to refuse benign requests
          that superficially resemble harmful ones.
    \item \textbf{Reward hacking:} The model may learn outputs that satisfy the DPO
          loss without being genuinely helpful or harmless.
    \item \textbf{Adversarial misuse:} An aligned model can still be misled via prompt
          injection or jailbreaks not represented in the training distribution.
\end{itemize}

% ─────────────────────────────────────────────────────────────
\section{GitHub and Reproducibility Status}

\textbf{Repository:} \url{https://github.com/ugurcavusoglu/ceng467-alignment-dpo}

\subsection*{Current Repository Structure}

\begin{lstlisting}
ceng467-alignment-dpo/
├── data/
│   └── prepare_dataset.py      # HH-RLHF download and preprocessing
├── scripts/
│   ├── train_sft.py            # SFT baseline (TinyLlama + LoRA)
│   ├── train_dpo.py            # DPO fine-tuning (TRL DPOTrainer)
│   └── run_eval.py             # Perplexity + BERTScore evaluation
├── results/
│   └── tables/
│       └── eval_results.json   # Evaluation output (all 3 models)
├── report/
│   └── progress_report.tex
├── requirements.txt
├── .gitignore
└── README.md
\end{lstlisting}

\subsection*{Reproducibility Instructions}

\begin{lstlisting}[language=bash]
git clone https://github.com/ugurcavusoglu/ceng467-alignment-dpo.git
cd ceng467-alignment-dpo
pip install transformers==4.44.2 trl==0.10.1 peft==0.12.0 \
            accelerate==0.34.0 datasets bert-score
python scripts/train_sft.py --subset 5000
python scripts/train_dpo.py --subset 3000
python scripts/run_eval.py --n_samples 100
\end{lstlisting}

\subsection*{Repository Status}

\begin{table}[h]
\centering
\begin{tabular}{ll}
\toprule
\textbf{Item} & \textbf{Status} \\
\midrule
GitHub repository                        & Public \\
\texttt{requirements.txt}                & Done \\
README with reproduction steps           & Done \\
\texttt{data/prepare\_dataset.py}        & Done \\
\texttt{scripts/train\_sft.py}           & Done \\
\texttt{scripts/train\_dpo.py}           & Done \\
\texttt{scripts/run\_eval.py}            & Done \\
\texttt{results/tables/eval\_results.json} & Done \\
Consistent commit history                & Active \\
\bottomrule
\end{tabular}
\caption{Repository status checklist}
\end{table}

% ─────────────────────────────────────────────────────────────
\section{Current Challenges and Next Steps}

\subsection*{Current Challenges}

\begin{itemize}[noitemsep]
    \item \textbf{GPU memory:} DPO requires loading both the policy model and the
          frozen reference model simultaneously, increasing memory pressure relative
          to SFT. TinyLlama with LoRA fits on a T4/A100, but longer sequences are
          constrained.
    \item \textbf{Session limits:} Colab/Kaggle session timeouts require a
          checkpointing strategy for longer training runs at the 50k+ scale.
    \item \textbf{TRL API instability:} TRL v1.3.0 introduced breaking API changes;
          we pinned to v0.10.1 for stability.
    \item \textbf{Limited training data:} Current results use 3--5k samples; scaling
          to 50k+ is needed for final evaluation and meaningful ablation comparisons.
\end{itemize}

\subsection*{Next Steps Before Final Submission}

\begin{enumerate}[noitemsep]
    \item Scale SFT and DPO training to 50k+ samples on GPU
    \item Run the full ablation study over $\beta$, learning rate, and LoRA rank
    \item Implement win-rate evaluation with a reward model
    \item Perform error analysis on 20--30 failure cases
    \item Write the final report in LNCS format (6--8 pages)
    \item Prepare live demo
\end{enumerate}

% ─────────────────────────────────────────────────────────────
\begin{thebibliography}{9}

\bibitem{rafailov2023}
Rafailov, R., Sharma, A., Mitchell, E., Ermon, S., Manning, C. D., \& Finn, C. (2023).
\textit{Direct Preference Optimization: Your Language Model is Secretly a Reward Model}.
arXiv:2305.18290.

\bibitem{bai2022}
Bai, Y., Jones, A., Ndousse, K., Askell, A., Chen, A., DasSarma, N., ... \& Kaplan, J. (2022).
\textit{Training a Helpful and Harmless Assistant with Reinforcement Learning from Human Feedback}.
arXiv:2204.05862.

\bibitem{ouyang2022}
Ouyang, L., Wu, J., Jiang, X., Almeida, D., Wainwright, C. L., Mishkin, P., ... \& Lowe, R. (2022).
\textit{Training language models to follow instructions with human feedback}.
NeurIPS 2022.

\bibitem{hu2021}
Hu, E. J., Shen, Y., Wallis, P., Allen-Zhu, Z., Li, Y., Wang, S., ... \& Chen, W. (2021).
\textit{LoRA: Low-Rank Adaptation of Large Language Models}.
arXiv:2106.09685.

\end{thebibliography}

\end{document}
